---
title: "CourseWork"
output: html_document
date: "2023-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

### Exploratory Analysis

```{r}
obs_data <- read.csv('Kittiwake_Observation_20594616.csv')
obs_data
```
```{r}
summary(obs_data)

dawn <- obs_data$dawn
noon <- obs_data$noon
afternoon <- obs_data$mid.afternoon
dusk <- obs_data$dusk

```
```{r}
print('Standard Deviation:')
paste('Dawn: ', sd(dawn))
paste('Noon: ', sd(noon))
paste('Afternoon: ', sd(afternoon))
paste('Dusk: ', sd(dusk))
```

```{r}
boxplot(obs_data, xlab = "Time Periods", ylab = "Sightings")
```
```{r}

# Plot histograms and overlay normal distribution curves
hist(dawn, freq = FALSE, main = "Distribution of Dawn Observations", xlim = range(dawn))
curve(dnorm(x, mean(dawn), sd(dawn)), col = "blue", add = TRUE)

hist(noon, freq = FALSE, main = "Distribution of Noon Observations", xlim = range(noon))
curve(dnorm(x, mean(noon), sd(noon)), col = "red", add = TRUE)

hist(afternoon, freq = FALSE, main = "Distribution of Afternoon Observations", xlim = range(afternoon))
curve(dnorm(x, mean(afternoon), sd(afternoon)), col = "green", add = TRUE)

hist(dusk, freq = FALSE, main = "Distribution of Dusk Observations", xlim = range(dusk))
curve(dnorm(x, mean(dusk), sd(dusk)), col = "purple", add = TRUE)


```
```{r}
# Combine all data into a single list for easy iteration
observation_data <- list(dawn, noon, afternoon, dusk)
colors <- c("blue", "red", "green", "purple")
names <- c("Dawn", "Noon", "Afternoon", "Dusk")

# Create an empty plot to prepare for overlaying histograms and curves
plot(1, type = "n", xlim = c(min(unlist(observation_data)), max(unlist(observation_data))), ylim = c(0, 0.15), 
     main = "Comparison Distributions with Normal Curves", xlab = "Observations", ylab = "Density")

# Loop through each observation data and overlay histograms with normal distribution curves
for (i in 1:length(observation_data)) {
  curve(dnorm(x, mean(observation_data[[i]]), sd(observation_data[[i]])), col = colors[i], add = TRUE)
}

# Add legend for better interpretation
legend("topright", legend = names, fill = colors)

```

### Confidence interval

```{r}
n <- length(obs_data$dusk)
sampmean<-mean(obs_data$dusk) # calculate the sample mean
sampvar<-var(obs_data$dusk) # calculate the sample variance
lower<-sampmean-qt(0.995,n-1)*sqrt(sampvar/n) # calculate the lower limit of the confidence interval from the formula
upper<-sampmean+qt(0.995,n-1)*sqrt(sampvar/n) # calculate the upper limit of the confidence interval from the formula
lower
upper
```

```{r}
dusk <- obs_data$dusk
result <- t.test(dusk, conf.level = 0.99)
result$conf.int
```

```{r}
result$estimate
```

## Task 2

```{r}
hist_data <- read.csv('Kittiwake_Historical_20594616.csv')
hist_data
```
```{r}
summary(hist_data[2:7])
```

```{r}
transposed_data <- t(hist_data[-1])
colnames(transposed_data) <- hist_data$X

library(ggplot2)

sites <- rownames(transposed_data)
plot_data <- data.frame(
  Year = rep(hist_data$X, times = ncol(hist_data) - 1),
  Kittiwake_Numbers = as.vector(as.matrix(hist_data[-1])),
  Site = rep(sites, each = length(hist_data$X))
)

ggplot(plot_data, aes(x = Year, y = Kittiwake_Numbers, color = Site)) +
  geom_line() +
  geom_point() +
  labs(title = "Kittiwake Numbers Over Time by Site", x = "Year", y = "Kittiwake Numbers")

```
```{r}
# Fit a linear regression model for each site
linear_models <- lapply(2:ncol(hist_data), function(i) lm(hist_data[, i] ~ X, data = hist_data))

# Get the slopes from the linear models
slopes <- sapply(linear_models, function(model) coef(model)[2])
slopes
```

### Independence of the decline and Site

```{r}
observed <- as.matrix(hist_data)[,-1]

yearly_count = rowSums(observed)
site_count = colSums(observed)
N <- sum(observed)

expected <- yearly_count%*%t(site_count)/N

expected
```

```{r}
chi_test<-sum((observed-expected)^2/expected)
chi_test
```
```{r}
nrows <- length(hist_data$X)
ncols <- length(hist_data[1,]) - 1
deg_freedom <- (ncols - 1) * (nrows - 1)
pvalue <- 1-pchisq(chi_test,deg_freedom)
pvalue
```
```{r}
chisq.test(observed)
```
### Prediction

```{r}
lm_model_site_A <- lm(Site.A ~ X, data = hist_data)
summary(lm_model_site_A)
```
```{r}
predicted_value <- predict(lm_model_site_A, newdata = data.frame(X = 2014))
predicted_value
```

## Task 3

```{r}
meas_data <- read.csv('Kittiwake_Measurement_20594616.csv')
meas_data
```
```{r}
summary(meas_data)
```
### Visual Summary

```{r}

library(ggplot2)

ggplot(meas_data, aes(x = Sub.species, y = Weight, fill = Sub.species)) +
  geom_boxplot() +
  labs(x = "Sub-species", y = "Weight") +
  theme_minimal()

ggplot(meas_data, aes(x = Sub.species, y = Wingspan, fill = Sub.species)) +
  geom_boxplot() +
  labs(x = "Sub-species", y = "Wingspan") +
  theme_minimal()

ggplot(meas_data, aes(x = Sub.species, y = Culmen, fill = Sub.species)) +
  geom_boxplot() +
  labs(x = "Sub-species", y = "Culmen") +
  theme_minimal()
```
```{r}
blacklegged <- subset(meas_data, Sub.species == 'Black-legged')
redlegged <- subset(meas_data, Sub.species == 'Red-legged')

summary(blacklegged[,-1])
summary(redlegged[,-1])
```


### Are Wingspan and Culmen Length independent for each specie

```{r}
plot(blacklegged$Wingspan,blacklegged$Culmen,xlab="Wingspan",ylab="Culmen Length", main = "Black-Legged Kittiwakes")
plot(redlegged$Wingspan,redlegged$Culmen,xlab="Wingspan",ylab="Culmen Length", main = "Red-Legged Kittiwakes")
```
```{r}
cor.test(blacklegged$Wingspan, blacklegged$Culmen)
```

```{r}
cor.test(redlegged$Wingspan, redlegged$Culmen)
```

### Comparison of Weights of both species

```{r}
t.test(Weight ~ Sub.species, data = meas_data)
```
### Difference between the 2 subspecies
```{r}
t.test(Weight ~ Sub.species, data = meas_data)
t.test(Wingspan ~ Sub.species, data = meas_data)
t.test(Culmen ~ Sub.species, data = meas_data)
```

```{r}
overall_test <- t.test(Weight + Wingspan + Culmen ~ Sub.species, data = meas_data)
overall_test
```

## Task 4

```{r}
loc_data <- read.csv('Kittiwake_Location_20594616.csv')
loc_data
```

```{r}
plot(loc_data)
```


## Predict the number of breeding pairs

```{r}
model1 <- lm(Breeding.pairs ~ ., data = loc_data)
summary(model1)
```
```{r}
model1 <- step(model1)
```
```{r}
summary(model1)
```
```{r}
hist(model1$residuals, main = "Histogram of Model Residuals", freq = FALSE)
curve(dnorm(x, mean(model1$residuals), sd(model1$residuals)), add = TRUE)
```



```{r}
plot(model1$fitted.values,model1$residuals,xlab="Fitted values",ylab="Residuals",main="Fitted Values Vs Residuals")
abline(h=0,lty=2)
```

## Linear Model for log of the number of breeding pairs
```{r}
model2 <- lm(log(Breeding.pairs) ~ ., data = loc_data)
summary(model2)
```
```{r}
model2 <- step(model2)
```
```{r}
summary(model2)
```
```{r}
hist(model2$residuals, main = "Histogram of Model Residuals", freq = FALSE)
curve(dnorm(x, mean(model2$residuals), sd(model2$residuals)), add = TRUE)
```



```{r}
plot(model2$fitted.values,model2$residuals,xlab="Fitted values",ylab="Residuals",main="Fitted Values Vs Residuals")
abline(h=0,lty=2)
```
### Most appropriate linear model

```{r}
AIC(model1)
AIC(model2)
```
### Model Fit

```{r}
plot(loc_data$sandeel, loc_data$Breeding.pairs, xlab = "Sandeel", ylab = "Breeding Pairs")
plot(loc_data$cliff.height, loc_data$Breeding.pairs, xlab = "Cliff Height", ylab = "Breeding Pairs")
plot(loc_data$sandeel, log(loc_data$Breeding.pairs), xlab = "Sandeel", ylab = "log(Breeding Pairs)")
plot(loc_data$cliff.height, log(loc_data$Breeding.pairs), xlab = "Cliff Height", ylab = "log(Breeding Pairs)")
```


### 80% confidence interval 
coastal direction = West, sandeel concentration = 2.93, mean summer temperature = 27.7 and cliff height (log) = 3.55.

```{r}
log_prediction <- predict(model2,newdata=data.frame(Coast.direction = 'West', sandeel = 2.93, Summer.temp = 27.7, cliff.height = 3.55),interval="confidence",level=0.80)
prediction <- exp(log_prediction)
log_prediction
prediction
```


